---
title: "Introduction to Statistical Analysis"
author: "Nancy Carmona & Rochelle-Jan Reyes"
date: "2024-5-10"
output: pdf_document
---

```{r set_knitr_options, echo=FALSE, message=FALSE, warning=FALSE}
suppressMessages(library(knitr))
opts_chunk$set(tidy=FALSE, cache=TRUE, echo=TRUE, message=FALSE)
```


# **Session 6 - Introduction to Statistical Analysis in R**

In this session you will learn:

* Utilize base R functions and packages for statistical analyses.
* Implement different types of statistical analyses methods, including simple analytics and comparative tests.
* Complete a statistical analysis in a real-world context through examples and demos.


## [***PPT Materials***]{.underline}

Here is the link to the Session 6 PPT. Feel free to use it as a reference when going through this RMD file: <https://docs.google.com/presentation/d/1dRIoI1cY5FXBbT89_Q4IsfuMQlqnlm34hY7UsrZ4NI8/edit?usp=sharing>

## [***R Functions***]{.underline}

Here is the link to the R Function Dictionary, with all of the functions you will need for this session under the Session 5 header: <https://docs.google.com/document/d/1z6agvbE-X70prlIdyFpU4-WkyI0IGI0WL5meEaQYz3o/edit?usp=sharing>


## A) Statistis Review

"Health statistics are numbers that summarize information related to health." (National Library of Medicine)

* Understand risk factors for communities
* Track and monitor health events
* See impacts of policy changes
* Assess the quality and safety of health care
* Used for evidence-informed policy making and evidence-based medicine

* Statistics is a branch of mathematics that deals with the manipulation, analysis, and interpretation of data (Merriam-Webster)

* Used in many different fields, including public health, epidemiology, and social sciences among other fields

Types of Statistical Studies
* Descriptive statistics
* Inferential statistics


## [***Activity 1 -  Descriptive Statistics in Base R***]{.underline} 

First, as always, let's load in our data and libraries (if any).

```{r}
ncha <- read.csv("NCHA2021_Session3_5_Subset.csv")
```

Okay, now we can go through some descriptive statistics.

Remember that descriptive statistics [*describe*]{.underline} what the data is. Inferential statistics is used to [*infer (draw) conclusions*]{.underline} from the data.

Let's look at the dataset really quickly. How do we *view* the data?

```{r}

```

Now let's look at age. What *class* is the variable AGE?

ANSWER:

Time to use some functions we've learned in section A to describe the data.

### Mean, Median, and Mode

First, let's look at mean (average). Which function can you use for finding the average value in the variable AGE?

```{r}

```

*Note*: will need to include na.rm = TRUE.

Now, we can look at the median.

```{r}

```

Lastly, we can look at mode. Normally, functions of what you need are named simply, like mean = mean() and median = median(). Mode is one of the few that aren't. You can find the mode (most of the same number of that variable) by using the *table()* function.

```{r}
table(ncha$AGE)
```

Is the mean, median, and mode the same number? If it's not the same number, then the data may be skewed! We can visualize this by using the base R plotting function, *hist()*.

```{r}

```

Which way is it skewed?

### Standard Deviation and Variance

Standard deviation (SD): how dispersed the data is in relation to the mean

```{r}

```

Variance: expected value of the squared deviation from the mean

```{r}

```

Great job! That's it for Activity 1 in Session 6.


## B) Statistical Analyses in R

_Key Steps for your Analysis_ 
* Set up analysis code
* Exploratory Analysis
* Data Wrangling
* Conduct descriptive analysis
* Conduct inferential analysis
* Document your analysis!


## C) Demo & Activity: Applying Descriptive Statistics

Install & Load *tidyverse* package

```{r}

# load the package "tidyverse" to wrangle data

library(tidyverse)

```

Explore the NCHA dataset! What does your data structure look like? 

```{r}

# View the structure of the dataset after it has been imported and saved to your working environment

glimpse(ncha)

```

Explore the NCHA dataset! What does your data structure look like? 


# Tidyverse also has functions which we can use to "summarize" our data 

```{r}

# we can use statistical functions and combine them together 

# summarise is a tidyverse functions 

# we can save our table in an object

ncha_summary_table <- ncha %>% 
  
  # group by and calculate summary statistics
  summarise(Min = min(MIN_MOD_PA, na.rm = TRUE), 
            Max = max(MIN_MOD_PA, na.rm = TRUE), 
            Mean = mean(MIN_MOD_PA, na.rm = TRUE),
            Median = median(MIN_MOD_PA, na.rm = TRUE), 
            missing = sum(is.na(MIN_MOD_PA)), 
            .groups = "drop") %>% 
  
  # round by mutating on a predicate function
  mutate_if(is.numeric, round, 3)

```


## D) Demo & Activity: Applying Inferential Statistics

Now, let's say I am working on a project examining if there is a relationship between cannabis and alcohol use among college students at SFSU. I can use data from the American College Health Association – National College Health Assessment (ACHA-NCHA) which we used in Session 5. 

I have two variables of interest: 

_1_"Substances ever used - Alcoholic beverages (beer, wine, liquor, etc.)"

_2_"Substances ever used - Cannabis (marijuana, weed, hash, edibles, vaped cannabis, etc.) [nonmedical use]"


For categorical variables we can use "contingency tables" as part of the descriptive statistics step. I will create data table with our two variables of interest. 

I will need this table to use base R to visualize the data after we conduct a chi-squared test. 

```{r}

# Create a contingency table with alcohol ever and cannabis ever         
alcohol_cannabis_tbl <-
  table(ncha$SUB_ANY_ALCOHOL,
        ncha$SUB_ANY_CANNABIS)
                 
print(alcohol_cannabis_tbl)

```


I can create data subset with our two variables of interest that I will use for our inferential statistics. 

```{r}
# create a data subset from the main data set

alcohol_cannabis_df  <-
  data.frame(ncha$SUB_ANY_ALCOHOL, ncha$SUB_ANY_CANNABIS)

# view your new data subset 

print(table(alcohol_cannabis_df))

```


Now let's conduct the chi-square test! 
 
```{r}

# conduct Chi-square test 

chisq.test(alcohol_cannabis_df$ncha.SUB_ANY_ALCOHOL,
           alcohol_cannabis_df$ncha.SUB_ANY_CANNABIS, 
           correct=FALSE)

```

* What is the p-value? 
* Are the result's statistically significant? 


We can also make a barplot to visualize the chi-square data 

```{r}

# plot of Chi-Square test using base R plotting

barplot(
  alcohol_cannabis_tbl,
  beside = TRUE,
  col = c("lightblue", "lightgreen"),
  main = "Alcohol Use vs. Cannabis Use among SFSU Students in 2021",
  xlab = "Cannabis Use",
  ylab = "Number of Students"
)
# add legend to bar plot 
legend(
  "center",
  title = "Alcohol Use",
  legend = rownames(alcohol_cannabis_tbl),
  fill = c("lightblue", "lightgreen")
)

```



### Quick peak: Making tables easily in R

```{r}

# install packages for creating table 1 (descriptive stats table)
install.packages("table1")

# load the library into 
library(table1)

```


```{r}

# We can create a "table 1" with your original dataset, typically we will include all of the variables in our primary analysis 

table1( ~ SUB_ANY_ALCOHOL + SUB_ANY_CANNABIS +
          MIN_MOD_PA +
          MIN_VIG_PA,
          dat = ncha)


```



# Session Information and Code

The next two chunks should be included in the appendix of every R Markdown so  that you document your session information, and code in the document. This supports the reproducibility of your work!

```{r session.info}
#-----session information: beginning of Appendix -----------
# This promotes reproducibility by documenting the version of R and every package
# you used.
sessionInfo()
```


```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE, include=TRUE}
#-----appendix------------
```

