---
title: "R Data Types & Language Basics"
author: "Nancy Carmona"
date: "2024-03-21"
output: pdf_document
---

```{r set_knitr_options, echo=FALSE, message=FALSE, warning=FALSE}
suppressMessages(library(knitr))
opts_chunk$set(tidy=FALSE, cache=TRUE, echo=TRUE, message=FALSE)
```

# Session 2. R Language Basics - Learning Objectives

By the end of this session you will learn:

* The basics of R "data objects" and "classes"
* The basic "primitive" data types in R
* How to inspect the properties of data objects
* How to create categorical variables in R
* How to convert data objects to other types
* How to access, assign and display data in R
* How to call functions
* How to use vectorized operations
* How to store, load, and remove data
* How to make comments in R code
* What is unusual about R
* The overall structure of the R language
* The basic "higher order" data structures in R


## A) Data types & Classes 

### What are Data Objects?

R is an *object oriented* language - everything in R is an *object*. What 
this means is that all of your data and code structures are stored in your 
computer's memory using the same framework. 

There are a core set of basic object types and classes built into R. All other
objects are built upon these.

*Data objects* are those structures which you use to store data. Objects can 
also store functions, packages, connections, and other structures intended for 
other uses. 

Like other objects, data objects have a *type* and a *class*. The type defines
*how* the data are stored, while the class defines *what* data are contained 
within an object.


### Accessing Data

If you just enter a value at the R command prompt, R will print out the value 
to your screen:

```{r}
68.1
```

Or you can call the `print()` function instead (to be explicit about it):

```{r}
print('A')
```


### Data Types and Classes

The most basic data types are `numeric` (`double` and `integer`), `complex`, 
`logical` (Boolean), and `character` (string).

```{r}
typeof(1)       # "double"    -- double precision floating point number
typeof(1L)      # "integer"   -- "L" after one or more digits makes it an integer
typeof(TRUE)    # "logical"   -- either TRUE or FALSE
```

### Data Types and Classes: Character

Characters are entered with quotes around them (single or double quotes).

```{r}
typeof('abcd')  # "character" -- a non-numeric "string" of text characters
```

If you omit the quotes of an alphanumeric string, R will assume you are 
referring to a variable name.

```{r}
val <- 1234
typeof(val)
```

### Data Types and Classes

You can show the "higher order" (if any) type of a data object with `class()`.

```{r}
typeof(1234)        # double  (the more basic "primitive" type)
class(1234)         # numeric (the more general "higher order" type)
typeof('abcd')      # character
class('abcd')       # character
```


### Dates and Times

There are special classes for dates and time to allow for date/time arithmetic 
and formatting.

```{r}
# Class: "Date"
my_date <- as.Date("2020-09-30", format = '%Y-%m-%d')

# Class: "POSIXct" (date, time, timezone, etc.) and "difftime"
my_datetime <- as.POSIXct("2020-09-30 12:00:", tz = "US/Pacific")
```

### Single-Value (Scalar) Data Types

Data types consisting of only a single value are also called "scalar" types.

Example    | Type      | Class             | Storage Mode
-----------|-----------|-------------------|-------------
1L         | integer   | integer           | numeric
1          | double    | numeric           | numeric
'1'        | character | character         | character
TRUE       | logical   | logical           | logical
Sys.Date() | double    | Date              | numeric 
Sys.time() | double    | POSIXct,  POSIXlt | numeric 

### Data Types and Classes: Factors

Factors are categorical variables. Let's create an example and explore its 
properties.

```{r}
country <- factor(c("usa", "canada", "mexico", "usa", "mexico"))
country
class(country)
```

This looks like a character variable with the additional attribute of "Levels", 
where the levels are the unique values. But how is this implemented?

### Factors

```{r}
typeof(country)
str(country)
attributes(country)
```

So factors in R are objects of class "factor" composed of an integer vector and 
a `$class` attribute and a `$levels` attribute containing a character vector.


# B) Creating Data 

### Variable assignment with `<-`

The `<-` (arrow) symbol is the primary assignment operator in R.

To store data in memory (in the environment) *assign* values to variables (using `<-`).

```{r}
number <- 68.1
number
```

This value will be stored in your computer's memory for the duration of the 
current session, or until it is modified or deleted by you, your code, or your
environment (execution context).

### Argument assignment with `=`

The `=` symbol is an additional assignment operator, most generally used within function calls.

The `=` symbol tells R that the expected argument (parameter) `x` should take 
the value of our `number` variable.

```{r}
print(x = number)
```

This assignment takes place within the function call and is not available 
outside of the function or after the function call has competed.

### Implicit (unnamed) function arguments

Since the `x` argument is the first parameter expected by the function, we do
not have to explicitly assign our value as a named argument, if we supply our
value as the first (or only) argument.

```{r}
print(number)
```

### Assignment operators

Just remember: 

* use `<-` for variable assignment (but you may also see `=` here too)

```{r, eval = FALSE}
number <- 68.1
```

* use `=` for argument assignment within function calls (almost never `<-`)

```{r, eval = FALSE}
print(x = number)
```

### C) Data Structure

### Data Structures: Vector

In R, a vector is a group of values or variables. You can create a vector from a 
collection of data values using the `c()` (combine) function.

```{r}
id <- c('A', 'B', 'C', 'D')
origin <- factor(c("usa", "canada", "mexico", "canada"))
height <- c(68.1, 69.4, 71.2, 68.9)
weight <- c(159.2, 162.3, 203.5, 181.3)
```

Here, we created three vectors, which we may describe as follows:

* `id` is a character vector containing alphabetic identifiers
* `origin` is a factor vector containing countries of origin
* `height` is a numeric vector containing heights in inches
* `weight` is a numeric vector containing weights in pounds

More technically speaking, a vector is an ordered collection of one or more 
individual values of the same data type. 

In other computer languages this data structure would be called an array or a 
list, but R uses those terms for other, more complex data structures.

### Data Structures: Matrix

A Matrix is a two-dimensional structure of values, all of the same data type. 
It can be constructed from a Vector, as supplied by the "data" argument for the 
`matrix()` function.

```{r}
mat <- matrix(data = 1:9, nrow = 3)
mat
```

### Data Structures: Data Frame

You can store a two dimensional collection of data (a "table") as a data frame. This
is the most common way to work with data in R. A data frame is often constructed
from one or more vectors, where each vector becomes a column.

```{r}
df <- data.frame(id, origin, height, weight)
df
```

The rows represent observations or cases and the columns represent variables.

We can see that our data frame, `df`, is a data frame by using the `class()` function.

```{r}
class(df)
```


### Data Structures: List 

Lists are objects that can contain elements of different types (i.e. elements  may be heterogeneous or "non-atomic"). This is a complicated but very powerful idea. We can put any object in a list.

```{r}
l <- list(x = 1:3, z = LETTERS[1:2], df = df, frml = formula(height~origin))
l
```

### Data Structures: List

For example, this list was created with several named objects: 

* vectors of different lengths (`x` and `z`) 
* a data frame (`df`) 
* a formula (`frml`)

We can even include things like graphics (plot objects).

Let's use `sapply()` to see the class of each list item.

```{r}
sapply(X = l, FUN = class)
```

### Data Structures: List

Because a list is non-atomic, it can contain other non-atomic objects. 

```{r}
is.atomic(l)
```

Let's use `sapply()` to see if each list item is atomic or not.

```{r}
sapply(X = l, FUN = is.atomic)
```

The two vectors are atomic data types because each item they store can only be an individual value, like a single number or character string. 

The data frame and formula are not atomic because they can store items containing multiple values.

### Viewing data

You can see data frames in a grid format by using the `View()` function.

```{r, eval = FALSE}
View(df)
```

This will open the data frame in a new window or tab and display it like a 
spreadsheet.

However, unlike a spreadsheet, you will not be able to manipulate
your data in this display.

### Accessing Data Elements: Indexing

Data are accessed though ["indexing."](https://cran.r-project.org/doc/manuals/R-lang.html#Indexing) Today, we'll learn two primary methods.

Using our matrix example, `mat`, you can access data in the matrix by position 
using single brackets, `[`. 

```{r}
# Display the 4th element (which in this case is 4).
mat[4]

# Display the element in the first row and second column.
mat[1, 2]

# Display all of the rows of the second column.
mat[ , 2] 
```

### Accessing Data Elements by Variable Names

You can access data in data frames based on their position using double brackets (`[[`), 
but also with variable names using the `$` operator. 

```{r, eval = FALSE}
# Using position 
df[[1]]
df[[3]]
 
# Using variable names
df[['id']]
df[['height']]

# Using variable names
df$id
df$height
```

What is an advantage to indexing by column name?

### Performing calculations

R is basically a calculator. You can use many built-in operators and functions.

```{r}
2*2
2/2
2^2
sqrt(2)
```

Additional math functions that are useful:
* min(), max(). Calculates the minimum and maximum of a numeric vector
* sum(). Calculates the sum of a numeric vector
* mean(). Calculates the mean of a numeric vector
* range(). Calculates the minimum and maximum values of a numeric vector



### Vectorized operations

R allows you to perform calculations on all items in a vector or all rows of a 
data frame at the same time. 

Operations that allow this are called *vectorized* operations. They may be 
much faster than the alternatives (e.g. "loops"). 

```{r}
# Add a new column for adult BMI calculated from the weight and height columns
df$bmi <- (df$weight / df$height ^ 2) * 703
df
```


### Removing data

You can remove a data object from your working (volatile) memory with `rm()`.

```{r}
rm(df)
```



# D) Commenting

### Code Comments in R

The `#` is used for comments. Everything on a line of code after the `#` will be ignored by R.

```{r}
# Assign a value to "val".
val <- 1234
val    # This is a bad name for a variable because it is not very descriptive.
```

It's usually better to put comments on their own line, as in the first example. 
If comments are added to a line of code, as in the second example, both comments and code can be harder to read.

It's common to indent the comment to line up with the code it's commenting.

It's also common for the comment to read like a spoken-language command 
(e.g., "Do this"), so that it's like a verbal command you are giving to 
the computer. Then the code implements what the comment says to do.





# E) Demo


Now let's create new variables that will be stored in our environment.

```{r}
# Creating variables for height 
height1 <- 1.6

# creating variables for weight 
weight1 <- 50
```

Instead of creating individual variables we can create vectors with multiple elements Let's consider we have anthropometric data (noninvasive quantitative measurements of the body) from four individuals. 



```{r}
# Creating vectors for height and weight 
# height is in meters
# weight in kilograms 

height <- c(1.6, 1.8, 2.0, 2.5)
weight <- c(50, 60, 64, 95)


```

In R, you create a vector with the combine function c(). You place the vector elements separated by a comma between the parentheses. This is an atomic vector with elements of the same type. 

We can create a data frame in R by passing the vectors into the data.frame() function. Now let's combine those vectors above into a data frame with columns for height and weight. 

```{r}
# Creating a data frame by joining vectors of variables  
health_df <- data.frame(height, weight)
health_df

```

Let's view different features of that new data frame we just created. 

```{r}
# str() Displays the structure of an R object

str(health_df)

# ncol() Returns the number of columns of a matrix or a data frame

ncol(health_df)

# length() Returns the number of items in an R object, such as a vector, a list, and a matrix.

length(health_df)

# dim() Returns the dimensions of the data frame (ie. number of rows and number of columns)

dim(health_df)

```


We can also create new variables empty and add them into an existing data frame. 

```{r}
# We can add this empty vector the data frame
# This is just a place holder for the BMI variable we will calculate
health_df$BMI <- NA

# We can print the data fame and see the empty column

print(health_df)
```

Let's calculate BMI using the height and weight data we created above. We will add this new column to the existing data frame.   

```{r}
# We calculate BMI using the vectors we created above 

# BMI equation for reference: BMI = Weight / (Height ^ 2)
# ^ is used for exponents 

health_df$BMI <- (health_df$weight) / (height^2)
 
# We can also View the data frame in another tab of the Source pane 
View(health_df)

```


We can also use a function to round the values to a specific number of significance digits or decimal places. 

```{r}
# We can round values to 3 decimal places
round(health_df, digits = 3)

# We can also round our values to 3 significant values
signif(health_df, digits = 3)
```


##### Some Helpful Hints... 

```{r}
# You can use the template to create a data frame

first_column <- c("value_1", "value_2", "value_3")
second_column <- c("value_A", "value_B", "value_C")

df <- data.frame(first_column, second_column)

print(df)
```

```{r}
# You can use the template to create a data frame

df <- data.frame(first_column = c("value_1", "value_2", "value_3"),
                 second_column = c("value_A", "value_B", "value_C")
                 )

print(df)
```




# F) Activity

Create your own Data! 

**Hello World** in R Programming

When we start to learn any programming languages we do follow a tradition to begin HelloWorld as our first basic program. Here we are going to learn that tradition.

For the HelloWorld program, we just need a print function. No need of any packages or main functions, just a simple print function is enough.


```{r}
print("HelloWorld")
```

print() is a function which is used to print the values on to the output screen. It also has arguments, we can use it if needed. For example, in the above program, the output is printed with quotes by default we can remove it if needed. We passed the argument named quote in print function which value is set to be false to remove the quotes.


#### Create and retrieve information about variables in R 

Double check that you do not have stored objects in your current session with the following command. This will list all objects that you have in your current R session.

```{r}
ls()
```


1) Create three vectors a, b, c with three integers each. 

```{r}


```


2) Check the data class and type for each vector. 

```{r}

```


3) Calculate the mean, median, and mode of each vector. 


```{r}


```


4) Combine the three vectors into a data frame. 


```{r}

```


5) Print the content of the data frame. 


```{r}

```



