---
title: "Session 5 - Data Visualization (Activity Only)"
author: "Rochelle-Jan Reyes"
date: "2024-05-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Session 5 - Introduction to Data Visualization & ggplot2**

## [***PPT Materials***]{.underline}

Here is the link to the Session 5 PPT. Feel free to use it as a reference when going through this RMD file: <https://docs.google.com/presentation/d/1f6Rf-Z_BJFmi6_P08fBiLW6MoTTbNMOjV9DGOUH2LGM/edit?usp=sharing>

## [***R Functions***]{.underline}

Here is the link to the R Function Dictionary, with all of the functions you will need for this session under the Session 5 header: <https://docs.google.com/document/d/1z6agvbE-X70prlIdyFpU4-WkyI0IGI0WL5meEaQYz3o/edit?usp=sharing>

## [***Activity***]{.underline}

First, make sure that you have the ggplot2 package AND the data loaded in to your R Session

```{r}
# load in the packages ggplot2 (for plotting) and RColorBrewer (for colors)
library(ggplot2)
#alternate option: library(tidyverse)

install.packages("RColorBrewer")
library(RColorBrewer)

# load in the data
ncha <- read.csv(file = "NCHA2021_Session3_5_Subset.csv", header = TRUE, na.strings=c("","NA"))

```

Next, let's view the data. How do we view the data?

```{r}

```

Now, let's check the class of each variable.

```{r}

```

Great! Now let's start plotting.

Age is a continuous variable. How can we plot age to see the ages of the participants in the survey in base R?

```{r}

```

How about in ggplot?

```{r}

```

Now, let's try to figure out whether there is a difference of minutes of moderate physical activity (continuous variable) among the different values of overall health (discrete variable) with a boxplot.

```{r}

```

Next, let's look at the age (continuous variable) of survey respondents and their minutes of vigorous physical activity (continuous variable) with the scatterplot function.

Note: in the plot type function, include the following: position = "jitter"

```{r}

```

Amazing! Okay, now let's start customizing. Select a plot that you've made, copy and paste that code into the code block below, and start customizing to your heart's desire!

Things you can do:

-   Add a fill or color variable inside the aes() function

-   Label your axes and titles (should do!)

-   Scale the x- or y-axes

-   Change the aesthetics with a theme preset or theme()

Keep running the code every once in awhile to check if your code works!

```{r}

```

Great job! You just learned how to plot data for visualization!!!
