---
title: "Session5_DataViz-ggplot2"
author: "Rochelle-Jan Reyes"
date: "2024-03-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Session 5! Today, we will be going over data visualization in R and the package ggplot2.

# **Session 5 - Introduction to Data Visualization & ggplot2**

## [***PPT Materials***]{.underline}

Here is the link to the Session 5 PPT. Feel free to use it as a reference when going through this RMD file: <https://docs.google.com/presentation/d/1f6Rf-Z_BJFmi6_P08fBiLW6MoTTbNMOjV9DGOUH2LGM/edit?usp=sharing>

## [***R Functions***]{.underline}

Here is the link to the R Function Dictionary, with all of the functions you will need for this session under the Session 5 header: <https://docs.google.com/document/d/1z6agvbE-X70prlIdyFpU4-WkyI0IGI0WL5meEaQYz3o/edit?usp=sharing>

## [***Session 5 Content***]{.underline}

### Set-Up

Let's load in the data that we will use in this session

```{r}
ncha <- read.csv(file = "NCHA2021_Session3_5_Subset.csv", header = TRUE, na.strings=c("","NA"))
```

### A. Data Visualization (Data Viz) in R

*Definition*: Representation of data through use of common graphics, such as plots, infographics, and animations

*Purpose*: communicate complex data relationships and data-driven insights in a way that is easy to understand

Data visualization can be done in R through base R functions or through many different packages, including the ggplot2 package in Tidyverse.

### B. Data: Discrete v. Continuous

**Discrete:** data type that is counted, with typically integer values

**Continuous**: data type that is measured, which can typically go to infinity and/or be values in between whole numbers

### C. Base R Plots

Base R provides an easy way to make simple plots. No packages are needed to carry out data viz! You can create different types of plots via base R functions, including barplots, histograms, box plots, and scatterplots.

#### General Plot Functions

```{r}
#Barplot Function
barplot(table(data$var1))
```

```{r}
#Histogram Function
hist(data$var1)

#Changing bin breaks
hist(data$var1, breaks = 10)
```

```{r}
#Boxplot Function
boxplot(data$var1, data$var2)
```

```{r}
#Scatterplot Function
plot(data$var1, data$var2)
```

#### *Code*: Dissecting parts of the function

```{r}
plot(data$var1, data$var2, #2 variables you want to visualize
     main = "Title", #Title of the plot
     sub = "Subtitle", #Subtitle of the plot
     xlab = "x-axis variable name", #Name of the x-axis
     ylab = "y-axis variable name", #Name of the y-axis
     xlim = c(0, 100), #Range to view for x-axis
     ylim = c(0, 100)) #Range to view for y-axis

```

### D. ggplot2 Plots

First, we need to load in the library

```{r}
library(ggplot2)
#OR you can load in the tidyverse package that contains ggplot2
#library(tidyverse)
```

#### What the code can look like

Simple code

```{r}
ggplot(ncha) + 
  geom_histogram(aes(x = AGE))
```

Complex code

```{r}
ggplot(ncha) + 
  geom_histogram(aes(x = AGE, fill = OVRL_HEALTH), bins = 20, col="white") +
  scale_x_continuous(n.breaks = 20) + 
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2") + 
  xlab("Age") +
  ylab("Number of Students") +
  ggtitle("The number of Students by Age and Overall Health") +
  theme_light()

```

#### Plot Types

There are many plot types. The type of plot you choose is dependent on the variables you'd like to explore (i.e., discrete or continuous?)

```{r}
#Bar Plot Function
geom_col(aes(x = x))
```

```{r}
# Histogram Function
geom_histogram(aes(x = x))
```

```{r}
# Boxplot Function
geom_boxplot(aes(x = x, y = y))
```

```{r}
# Scatterplot Function
geom_point(aes(x = x, y = y))
```

#### Code: Dissecting parts of the function

```{r}
#calling in the ggplot function
ggplot(ncha) +
  #select the plot type
  geom_boxplot(
    #use aes to call in the variables
    aes(x = OVRL_HEALTH, y = SUGAR_7D)) +
  #Name the x-axis
  xlab("Reported Overall Health") + 
  #Name the y-axis
  ylab("# of Sugar-Sweetened Beverages (last 7 days)") + 
  #Name the plot
  ggtitle("Sugar-Sweetened Beverages in the last 7 days and Reported Overall Health")

```

#### Fill Variable

The fill() function can allow you to have an additional, discrete variable to categorize your data in the figure.

```{r}
ggplot(data) +
  geom_col(aes(x = var1, fill = var2))
```

#### Scales

Scales in ggplot allow you to make changes to your x- and y- axes.

```{r}
#this is the base part of the plot 
p <- ggplot() + geom_point(aes())

# x-axis scales
p + scale_x_discrete(breaks = c(), labels = c())
p + scale_x_continuous(breaks = c(), labels = c())
p + scale_x_log10()

# y-axis scales
p + scale_y_discrete(breaks = c(), labels = c())
p + scale_y_discrete(breaks = c(), labels = c())
p + scale_y_log10()

# fill scales
p + scale_fill_brewer(labels = c(), palette = "nameofpalette")
```

#### Themes

Change the aesthetic of your plot(s)

```{r}
#to fully customize your plot, use this function after your base ggplot code
+ theme()

#what it would look like with the base ggplot code
ggplot() + geom_point(aes()) + 
  theme()
```

See all the things you can change in the theme function here: <https://www.rdocumentation.org/packages/ggplot2/versions/3.5.0/topics/theme> 

##### Preset Themes

```{r}
#light grey background
theme_grey()

#outlined part of the plot
theme_bw()

# inner grid lines are black
theme_linedraw()

# white background and light grey lines
theme_light()

#dark grey background
theme_dark()

#white background, no outline
theme_minimal()

# no inner gridlines
theme_classic()

# no axes titles, ticks, or inner gridlines
theme_void()

```

### E. Examples

See PPT for examples

### F. Activity

First, make sure that you have the ggplot2 package AND the data loaded in to your R Session

```{r}
# load in the packages ggplot2 (for plotting) and RColorBrewer (for colors)
library(ggplot2)
#alternate option: library(tidyverse)

install.packages("RColorBrewer")
library(RColorBrewer)

# load in the data
ncha <- read.csv(file = "NCHA2021_Session3_5_Subset.csv", header = TRUE, na.strings=c("","NA"))
```

Next, let's view the data. How do we view the data?

```{r}

```

Now, let's check the class of each variable.

```{r}

```

Great! Now let's start plotting.

Age is a continuous variable. How can we plot age to see the ages of the participants in the survey in base R?

```{r}

```

How about in ggplot?

```{r}

```

Now, let's try to figure out whether there is a difference of minutes of moderate physical activity (continuous variable) among the different values of overall health (discrete variable) with a boxplot.

```{r}

```

Next, let's look at the age (continuous variable) of survey respondents and their minutes of vigorous physical activity (continuous variable) with the scatterplot function.

Note: in the plot type function, include the following: position = "jitter"

```{r}

```

Amazing! Okay, now let's start customizing. Select a plot that you've made, copy and paste that code into the code block below, and start customizing to your heart's desire!

Things you can do:

-   Add a fill or color variable inside the aes() function

-   Label your axes and titles (should do!)

-   Scale the x- or y-axes

-   Change the aesthetics with a theme preset or theme()

Keep running the code every once in awhile to check if your code works!

```{r}

```

Great job! You just learned how to plot data for visualization!!!
