---
title: "Session3_Packages-Scripts_ActivityONLY"
author: "Rochelle-Jan Reyes"
date: "2024-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 3 - R Sessions, Packages & Scripts, and R File Types (Importing & Exporting) --- ACTIVITIES ONLY

## Functions Addressed

Installing Packages

```{r}
install.packages("")

```

Reading in the package to the library

```{r}
library()
```

Importing in data sets / data files

```{r}
# EXCEL or TXT files
df_txt <- read.table("TXTFile_Template.txt", header = TRUE)
df_exc <- read.table("EXCELFile_Template.xlsx", header = FALSE)

# CSV files
df_csv <- read.csv("CSVFile_Template.csv", header = TRUE)
```

Accessing Infromation from the dataset

```{r}
# Viewing the data frame
View() #note that it's a capital V

# View the internal structure of the data frame
str()

# Find the number of different values in a variable
table(df$variable)
```

Exporting data sets into files

```{r}
# TXT files
write.table(mtcars, file = "mtcars.txt", sep = "\t",
            row.names = TRUE, col.names = NA)

# CSV files
write.csv(mtcars, file = "mtcars.csv")
```

## *Activity 1:* Create an RStudio Project for the Workshop

**Instructions:** 

1.  Go to *File*

2.  Select *New Project*

3.  Select *New Directory*

4.  Select *New Project*

5.  Change name to \"r_onramp\"

6.  Change working directory to desired computer location

7.  Press button on \"Open in New Session\"

8.  Create that new project!

## *Activity 2*: Importing, Viewing, and Exporting

Now it's your turn to import and export data, with a little bit of functions we learned from last session! [Remember to write your code in the pre-loaded coding blocks]{.underline}.

#### Importing the Data

First, let's import the data set. *The dataset is downloadable from Module 3's Workshop Materials on Canvas. Hint: the dataset is in a CSV file form and we need to **read** it into the Session environment.*

```{r}
#insert the code here

```

#### Working with the Data

Great! Now you have the data in your environment! The next step of this activity is to view the data set.

```{r}
# view the data set

```

Questions:

```{r}
# you can use this space to answer the questions below
```

1.  How many variables are in the data set? *Hint: you can use either View() or str() to answer this question!*

2.  What two types/classes of data are present in the data set? *Hint: use the function str() to answer the questions.*

Next, we are going to look at the values in the variable OVRL_HEALTH. How do we view the number of different values in a variable?

```{r}
# Input code here

```

Questions:

1.  How many different values does the variable OVRL_HEALTH have?

#### Creating a data frame with the current data set

We can create a data frame with data from the current data set by using the function, data.frame(). Select two or more variables from the current data set and create a new data frame.

```{r}
# Insert code below

```

#### Exporting the Data

Great! Now, let's export that data frame you made as a .txt file. *Hint: use the write.table() function.*

```{r}
# Insert code below

```

# **Great job, you finished Session 3!**
