---
title: "Downloading SVI Data"
output: html_document
date: "2025-07-16"
editor_options: 
  markdown: 
    wrap: 72
---

```{r set.knitr.options, include= FALSE, message=FALSE, warning=FALSE}
#-----set knitr options-----

# clear environment
rm(list = ls())

# knitr options
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, echo = TRUE, message = FALSE)

# Load pacman, installing if needed and other packages as needed
if (!require("pacman")) { install.packages("pacman", repos = "http://cran.r-project.org") }

# Load other packages, installing as needed.
pacman::p_load(tidyverse, tigris, sf, tmap, leaflet, RColorBrewer)

# Set working directory 
setwd("~/PINC 2025/Bilaoen-PSP-2025/Climate Change Data")
```

# About SVI Data

The Social Vulnerability Index (SVI), developed by the CDC/ATSDR,
measures the resilience of communities when confronted by external
stresses such as natural disasters or disease outbreaks. It is based on
16 U.S. Census variables grouped into four themes: socioeconomic status,
household composition & disability, minority status & language, and
housing type & transportation. The SVI is reported at the census tract
level and ranges from 0 (least vulnerable) to 1 (most vulnerable),
allowing for comparison across geographic areas.

In our analysis, we use the 2020 SVI to assess how extreme heat exposure
correlates with social vulnerability across San Francisco census tracts
from 2000 to 2020. By linking weekly heat metrics derived from Daymet
data to the SVI scores of each tract, we examine whether communities
with higher social vulnerability experience more frequent or intense
extreme heat events. This helps us identify environmental justice
patterns and prioritize areas for future public health interventions.

The variable *RPL_THEMES* represents the overall percentile rank of
social vulnerability for each census tract, based on the combination of
all four SVI themes: Socioeconomic status Household composition &
disability Minority status & language Housing type & transportation

## Downloading Data

Since no specific API exists, we can directly download from the website:
<https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html>

You can select Year: 2020, Geography: California, Geography Type: Census
Tracts.

```{r load.data, echo= FALSE}

# Step 0: Load SVI and pre-processed yearly heat wave data 
svi_data <- read_csv("CA_SVI.csv")
# View(svi_data)

heat_wave_summary <- read_csv("heat_wave_summary.csv")
# View(heat_wave_summary)

```




```{r, echo = FALSE}

# Join with SVI Data

# since the svi_data has a FIPS code (as a character) with a leading 0, but the heat_wave_summary has site (as a double) without the leading 0, data wrangling is needed...
# we can fix this by converting site to character with leading zeroes

heat_wave_summary_site <- heat_wave_summary %>%
  mutate(site = as.character(sprintf("%011.0f", site)))


yearly_ehe_svi <- heat_wave_summary_site %>%
  left_join(svi_data, by = c("site" = "FIPS"))


# filter the SVI to only keep valid values 
yearly_ehe_svi <- yearly_ehe_svi %>%
  filter(RPL_THEMES >= 0 & RPL_THEMES <= 1)



# > dim(yearly_ehe_svi)
# [1] 5082  163

# > dim(yearly_ehe_svi_filtered)
# [1] 5019  163

# > length(unique(yearly_ehe_svi$site))
# [1] 242

# > length(unique(yearly_ehe_svi_filtered$site))
# [1] 239


```


```{r}

# aggregate data by site 

tract_summary <- yearly_ehe_svi %>%
  group_by(site, RPL_THEMES) %>%
  summarize(
    mean_hw_2day_85 = mean(hw_2day_85, na.rm = TRUE),
    mean_hw_3day_85 = mean(hw_3day_85, na.rm = TRUE),
    mean_hw_2day_97_5 = mean(hw_2day_97_5, na.rm = TRUE),
    mean_hw_3day_97_5 = mean(hw_3day_97_5, na.rm = TRUE),
    .groups = "drop"
  )


```


## Correlations

The commonly used correlations in data science are “Spearman and Pearson.” The difference between Spearman and Pearson is that  Spearman uses rank variables while Pearson uses interval or ratio variables. 

Use Pearson correlation for linear relationships with normally distributed data, and Spearman correlation for monotonic relationships, or when data is ordinal or not normally distributed. Pearson measures the strength of a linear association, while Spearman assesses the strength of a monotonic association, which can be linear or non-linear. 

Note: If the relationship between temperature and heatwave data is expected to be linear and normally distributed with no outliers, Pearson correlation is the standard choice. However, if the relationship is monotonic but not strictly linear, or if the data is non-normal or contains outliers, Spearman correlation provides a more robust and accurate measure of association. 

Let's examine correlations with the average number of heatwaves (by site, aka. census tract) and SVI. 

For a Pearson correlation between SVI and the average annual number of 3-day heatwaves (97.5th percentile) per tract from 2000–2020, we found a positive correlation between SVI and heatwave exposure.

A Pearson correlation analysis revealed a statistically significant positive association between social vulnerability (RPL_THEMES) and average annual 3-day heatwave events (97.5th percentile) across census tracts in San Francisco from 2000–2020 (r = 0.22, p < 0.001). This suggests that tracts with higher social vulnerability tend to experience more frequent extreme heat events, though the strength of the relationship is modest.

```{r}

# Spearman correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_2day_85,
         method = "spearman")

# Pearson correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_2day_85,
         method = "pearson")

#  p-value = 0.02176
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.27018047 -0.02188942
# sample estimates:
#        cor 
# -0.1483723 

# Spearman correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_3day_85,
         method = "spearman")

# Pearson correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_3day_85,
         method = "pearson")

# p-value = 0.0004577
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.3420771 -0.1009326
# sample estimates:
#        cor 
# -0.2249466 

# Spearman correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_2day_97_5,
         method = "spearman")

# Pearson correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_2day_97_5,
         method = "pearson")


# p-value = 0.01366
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  -0.28056003 -0.03311216
# sample estimates:
#        cor 
# -0.1593377 


# Spearman correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_3day_97_5,
         method = "spearman")

# Pearson correlation
cor.test(tract_summary$RPL_THEMES,
         tract_summary$mean_hw_3day_97_5,
         method = "pearson")

# p-value = 0.0007632
# alternative hypothesis: true correlation is not equal to 0
# 95 percent confidence interval:
#  0.0918933 0.3339956
# sample estimates:
#       cor 
# 0.2162665 

# cor = 0.2163
# This means there is a positive correlation between SVI and heatwave exposure.
# 
# The strength of the relationship is weak to moderate — tracts with higher social vulnerability tend to experience more extreme heat events on average.
```

Scatterplots of SVI by Average annual number of heatwaves. 

```{r}

ggplot(tract_summary, aes(x = RPL_THEMES, y = mean_hw_2day_85)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Social Vulnerability Index",
    y = "Avg. Annual 2-Day Heatwaves (85th %)",
    title = "SVI vs. Heatwave Exposure by Tract"
  ) +
  theme_minimal()


```


```{r}
ggplot(tract_summary, aes(x = RPL_THEMES, y = mean_hw_3day_85)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Social Vulnerability Index",
    y = "Avg. Annual 3-Day Heatwaves (85th %)",
    title = "SVI vs. Heatwave Exposure by Tract"
  ) +
  theme_minimal()

```


```{r}

ggplot(tract_summary, aes(x = RPL_THEMES, y = mean_hw_2day_97_5)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Social Vulnerability Index",
    y = "Avg. Annual 2-Day Heatwaves (97.5th %)",
    title = "SVI vs. Heatwave Exposure by Tract"
  ) +
  theme_minimal()


```


```{r}

# scatterplot of SVI and average number of Annual 3-Day Heatwaves 

ggplot(tract_summary, aes(x = RPL_THEMES, y = mean_hw_3day_97_5)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    x = "Social Vulnerability Index",
    y = "Avg. Annual 3-Day Heatwaves (97.5th %)",
    title = "SVI vs. Heatwave Exposure by Tract"
  ) +
  theme_minimal()

```




```{r}

# Create SVI Quartiles 

tract_summary <- tract_summary %>%
  mutate(
    svi_quartile = ntile(RPL_THEMES, 4),     # Q1 = least vulnerable, Q4 = most
    svi_quintiles = ntile(RPL_THEMES, 5)    # Q1 = least vulnerable, Q5 = most
  )

```




## Plotting SVI and Average number of Annual Heatwaves

Let's look at a boxplot with annual EHE by SVI quartile. 

```{r}

ggplot(tract_summary, aes(x = svi_quartile, y = mean_hw_3day_97_5, color = factor(svi_quartile))) +
  geom_boxplot(size = 1.2) +
  scale_fill_brewer(palette = "YlOrRd", name = "SVI Quartile") +
  labs(
    title = "Avg. Annual 3-Day Heatwaves (97.5th %) by Social Vulnerability Index (SVI)",
    x = "SVI Quartiles",
    y = "Mean Heat Waves per Tract"
  ) +
  theme_minimal()


```


We can also use a time series plot comparing heatwave trends between the lowest and highest SVI tracts or neighborhoods to visually contrast vulnerability and exposure over time.


```{r}

library(dplyr)

ehe_svi_grouped <- yearly_ehe_svi %>%
  mutate(
    svi_group = case_when(
      RPL_THEMES >= quantile(RPL_THEMES, 0.90, na.rm = TRUE) ~ "High SVI",
      RPL_THEMES <= quantile(RPL_THEMES, 0.10, na.rm = TRUE) ~ "Low SVI",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(svi_group))

library(tidyr)

ehe_long <- ehe_svi_grouped %>%
  pivot_longer(
    cols = starts_with("hw_"),
    names_to = "definition",
    values_to = "event_count"
  )

heat_summary_long <- ehe_long %>%
  group_by(svi_group, year, definition) %>%
  summarise(mean_events = mean(event_count, na.rm = TRUE), .groups = "drop")

heat_summary_long <- heat_summary_long %>%
  mutate(definition = recode(definition,
    "hw_2day_85"   = "2-Day ≥ 85th",
    "hw_3day_85"   = "3-Day ≥ 85th",
    "hw_2day_97_5" = "2-Day ≥ 97.5th",
    "hw_3day_97_5" = "3-Day ≥ 97.5th"
  ))

library(ggplot2)

ggplot(heat_summary_long, aes(x = year, y = mean_events, color = svi_group)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~ definition, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c("High SVI" = "#d73027", "Low SVI" = "#1a759f")) +
  labs(
    title = "Heatwave Trends by SVI Group and Definition",
    subtitle = "Comparing High and Low Social Vulnerability Areas (2000–2020)",
    x = "Year",
    y = "Avg. Annual Heatwave Events",
    color = "SVI Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )


```



## Spatial Analysis 

### San Francisco Neighborhood Level Maps
```{r}

sf_neighborhoods <- read_csv("Analysis_Neighborhoods.csv")


# aggregate data by site 

annual_summary <- heat_wave_summary_site %>%
  group_by(site) %>%
  summarize(
    mean_hw_2day_85 = mean(hw_2day_85, na.rm = TRUE),
    mean_hw_3day_85 = mean(hw_3day_85, na.rm = TRUE),
    mean_hw_2day_97_5 = mean(hw_2day_97_5, na.rm = TRUE),
    mean_hw_3day_97_5 = mean(hw_3day_97_5, na.rm = TRUE),
    .groups = "drop"
  )



# Join with EHE + Neighborhood Data

neighborhood_data <- sf_neighborhoods %>%
  left_join(annual_summary, by = c("geoid" = "site"))

# add geometry 
# Convert to sf using WKT from 'the_geom' column
neighborhood_sf <- neighborhood_data %>%
  st_as_sf(wkt = "the_geom", crs = 4326) 


# remove The Farallones

neighborhood_sf <- neighborhood_sf %>%  filter(geoid != "06075980401")



# Heatwaves 
ggplot(neighborhood_sf) +
  geom_sf(aes(fill = mean_hw_2day_85), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 2-Day Heatwave Events (85th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


# aggregate data by neighborhoods_analysis_boundaries 

neighborhood_sf_avg <- neighborhood_sf %>%
  group_by(neighborhoods_analysis_boundaries) %>%
  summarize(
    mean_hw_2day_85 = mean(mean_hw_2day_85, na.rm = TRUE),
    mean_hw_3day_85 = mean(mean_hw_3day_85, na.rm = TRUE),
    mean_hw_2day_97_5 = mean(mean_hw_2day_97_5, na.rm = TRUE),
    mean_hw_3day_97_5 = mean(mean_hw_3day_97_5, na.rm = TRUE),
    .groups = "drop"
  )


# mapping at the neighborhood level 

ggplot(neighborhood_sf_avg) +
  geom_sf(aes(fill = mean_hw_3day_85), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 3-Day Heatwave Events (85th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



ggplot(neighborhood_sf_avg) +
  geom_sf(aes(fill = mean_hw_2day_97_5), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 2-Day Heatwave Events (97.5th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

ggplot(neighborhood_sf_avg) +
  geom_sf(aes(fill = mean_hw_3day_97_5), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 3-Day Heatwave Events (97.5th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```




```{r}
tmap_mode("plot")  # Use "view" for interactive mode

tm_shape(neighborhood_sf) +
  tm_fill("mean_hw_2day_85",
          title = "Mean HW (85th)",
          palette = "plasma",
          colorNA = "lightgray",
          textNA = "No data") +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(
    title = "Mean 2-Day Heatwave Events (85th Percentile)",
    frame = FALSE,     # Removes map frame
    legend.outside = TRUE,
    legend.title.size = 1.1,
    legend.text.size = 0.9,
    bg.color = "white"
  )

```


### Census Tract Level Maps



```{r}


# Heatwaves 
ggplot(neighborhood_sf) +
  geom_sf(aes(fill = mean_hw_2day_85), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 2-Day Heatwave Events (85th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplot(neighborhood_sf) +
  geom_sf(aes(fill = mean_hw_3day_85), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 3-Day Heatwave Events (85th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplot(neighborhood_sf) +
  geom_sf(aes(fill = mean_hw_2day_97_5), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 2-Day Heatwave Events (97.5th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


ggplot(neighborhood_sf) +
  geom_sf(aes(fill = mean_hw_3day_97_5), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Mean 3-Day Heatwave Events (97.5th Percentile)",
    fill = "Average Annual Heatwaves"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


# Pivot to long format for faceting
sf_mean_long <- neighborhood_sf %>%
  pivot_longer(
    cols = starts_with("mean_hw_"),
    names_to = "definition",
    values_to = "value"
  ) %>%
  mutate(definition = recode(definition,
    "mean_hw_2day_85"   = "2-Day ≥ 85th",
    "mean_hw_3day_85"   = "3-Day ≥ 85th",
    "mean_hw_2day_97_5" = "2-Day ≥ 97.5th",
    "mean_hw_3day_97_5" = "3-Day ≥ 97.5th"
  ))


ggplot(sf_mean_long) +
  geom_sf(aes(fill = value), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  facet_wrap(~ definition, ncol = 2) +
  labs(
    title = "Average Annual Heatwave Events by Definition (2000–2020)",
    fill = "Avg. Events"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


```




```{r, message = FALSE}

# Get the San Francisco County Boundary
sf_boundary <- counties(state = "CA", year = 2020, class = "sf") %>%
  filter(NAME == "San Francisco") %>%
  st_transform(4326)  # ensure consistent CRS

# Download Tracts for San Francisco
tracts_sf <- tracts(state = "CA", county = "San Francisco", year = 2020, class = "sf") %>%
  st_transform(4326)

tracts_sf <- tracts_sf %>%
  filter(ALAND > 0)  # Keeps only tracts with land area


tracts_sf_clean <- tracts_sf %>%
  filter(st_coordinates(st_centroid(geometry))[,1] > -123)  # Removes far west tracts


# Join with EHE + SVI Data

tracts_sf_data <- tracts_sf_clean %>%
  left_join(tract_summary, by = c("GEOID" = "site"))

```


Mapping the average number of heatwaves for each census tract over 2000-2020.

```{r}

tm_shape(tracts_sf_data) +
  tm_fill("mean_hw_3day_97_5", palette = "Reds", style = "quantile") +
  tm_borders() +
  tm_layout(title = "Annual Average Heatwave Exposure by Tract")


```

Mapping SVI. 

```{r}

tm_shape(tracts_sf_data) +
  tm_fill("RPL_THEMES", palette = "Blues", style = "quantile",
          title = "Social Vulnerability (SVI)") +
  tm_borders(alpha = 0.5) +
  tm_layout(title = "San Francisco Social Vulnerability by Census Tract")


```


Using leaflet we can also make an interactive map. 

```{r}
library(leaflet)
library(RColorBrewer)

# Heat palette
pal_hw <- colorQuantile("YlOrRd", domain = tracts_sf_data$mean_hw_3day_97_5, n = 5)

leaflet(tracts_sf_data) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal_hw(mean_hw_3day_97_5),
    color = "#FFFFFF", weight = 1, fillOpacity = 0.7,
    label = ~paste0("Tract: ", GEOID, "<br>Heatwaves: ", round(mean_hw_3day_97_5, 2)),
    highlightOptions = highlightOptions(weight = 2, color = "#000", bringToFront = TRUE)
  ) %>%
  addLegend("bottomright", pal = pal_hw, values = ~mean_hw_3day_97_5,
            title = "Avg. Annual 3-Day Heatwaves")

```


### Bivariate maps 

We can also create a bivariate map of SVI and Average Number of Heatwave days. 


```{r}

library(tibble)
library(ggplot2)
library(patchwork)
library(sf) 

# add svi to neighborhood_sf

neighborhood_sf_svi <- neighborhood_sf %>%
  left_join(svi_data, by = c("geoid" = "FIPS"))

neighborhood_bivar_all <- neighborhood_sf_svi %>%
  pivot_longer(
    cols = starts_with("mean_hw_"),  # e.g., mean_hw_2day_85, etc.
    names_to = "definition",
    values_to = "heat_value"
  )

neighborhood_bivar_all <- neighborhood_bivar_all %>%
  group_by(definition) %>%
  mutate(
    svi_q = ntile(RPL_THEMES, 4),
    heat_q = ntile(heat_value, 4),
    bivar_class = paste0(svi_q, "-", heat_q)
  ) %>%
  ungroup()

neighborhood_bivar_all <- neighborhood_bivar_all %>%
  mutate(definition = recode(definition,
    "mean_hw_2day_85"   = "2-Day ≥ 85th",
    "mean_hw_3day_85"   = "3-Day ≥ 85th",
    "mean_hw_2day_97_5" = "2-Day ≥ 97.5th",
    "mean_hw_3day_97_5" = "3-Day ≥ 97.5th"
  )) %>% 
  filter(definition == "2-Day ≥ 97.5th") #only looking at one heatwave definition

# Use the same bivar_palette you used in the map
bivar_palette <- c(
  "1-1" = "#f2f0f7", "1-2" = "#cbc9e2", "1-3" = "#9e9ac8", "1-4" = "#756bb1",
  "2-1" = "#e7d4e8", "2-2" = "#d1b9d8", "2-3" = "#b39dd0", "2-4" = "#8e6cbf",
  "3-1" = "#fbb4b9", "3-2" = "#f768a1", "3-3" = "#dd3497", "3-4" = "#ae017e",
  "4-1" = "#fcae91", "4-2" = "#fb6a4a", "4-3" = "#ef3b2c", "4-4" = "#99000d"
)

# Create 4x4 grid
bivar_legend_df <- expand.grid(
  svi_q = 1:4,
  heat_q = 1:4
) %>%
  mutate(
    bivar_class = paste0(svi_q, "-", heat_q),
    fill = bivar_class  # match to your map's fill values
  )

bivar_legend <- ggplot(bivar_legend_df, aes(x = heat_q, y = svi_q, fill = bivar_class)) +
  geom_tile() +
  scale_fill_manual(values = bivar_palette, guide = "none") +
  scale_x_continuous(
    breaks = 1:4,
    labels = c("Low", "", "", "High"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = 1:4,
    labels = c("Low", "", "", "High"),
    expand = c(0, 0)
  ) +
  labs(x = "Heatwave Exposure", y = "Social Vulnerability") +
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 8, face = "bold"),
    axis.text = element_text(size = 8),
    panel.grid = element_blank(),
    plot.margin = margin(2, 2, 2, 2)
  )

main_map <- ggplot(neighborhood_bivar_all) +
  geom_sf(aes(fill = bivar_class), color = "white", size = 0.2) +
  scale_fill_manual(values = bivar_palette, name = "SVI + Heat") +
# if you want to facet wrap facet_wrap(~ definition, ncol = 2) +
  labs(
    title = "Bivariate Maps of Social Vulnerability and Heat Exposure",
    subtitle = "Quartile Classification of SVI and Heatwave Metrics",
    caption = "Dark = High SVI and High Heat"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold", size = 12)
  )

final_plot <- main_map + 
  inset_element(bivar_legend, left = -0.5, bottom = 0.02, right = 0.25, top = 0.25)

print(final_plot)

```


```{r}
library(dplyr)


tract_sf_bivar <- tracts_sf_data %>%
    filter(!is.na(RPL_THEMES), !is.na(mean_hw_3day_97_5)) %>%
  mutate(
    svi_cat = ntile(RPL_THEMES, 3),               # 1 = low, 3 = high
    heat_cat = ntile(mean_hw_3day_97_5, 3),        # 1 = low, 3 = high
    bivar_class = paste0(svi_cat, "-", heat_cat)   # e.g., "3-3" = high SVI + high heat
  )

# choose a color palette - this can be changed if we don't like it! 

bivar_palette <- c(
  "1-1" = "#ffffcc", "2-1" = "#c2e699", "3-1" = "#78c679",
  "1-2" = "#fed976", "2-2" = "#feb24c", "3-2" = "#fd8d3c",
  "1-3" = "#fc4e2a", "2-3" = "#e31a1c", "3-3" = "#b10026"
)



library(tmap)

tmap_mode("plot")  # use "view" for interactive

tm_shape(tract_sf_bivar) +
  tm_fill("bivar_class",
          palette = bivar_palette,
          title = "SVI + Heatwave",
          labels = names(bivar_palette)) +
  tm_borders(alpha = 0.3) +
  tm_layout(title = "Bivariate Map: SVI and 3-Day Heatwave Exposure")

```


We can also make an interactive map with layer toggle!

```{r}

library(leaflet)
library(RColorBrewer)
library(sf)

# SVI palette (Blues)
pal_svi <- colorQuantile("Blues", domain = tracts_sf_data$RPL_THEMES, n = 5)

# Heatwave palette (Reds)
pal_hw <- colorQuantile("YlOrRd", domain = tracts_sf_data$mean_hw_3day_97_5, n = 5)

labels_svi <- paste0(
  "Tract: ", tracts_sf_data$GEOID, "<br>",
  "SVI: ", round(tracts_sf_data$RPL_THEMES, 2)
) %>% lapply(htmltools::HTML)

labels_hw <- paste0(
  "Tract: ", tracts_sf_data$GEOID, "<br>",
  "Avg. 3-Day Heatwaves: ", round(tracts_sf_data$mean_hw_3day_97_5, 2)
) %>% lapply(htmltools::HTML)

leaflet(tracts_sf_data) %>%
  addProviderTiles("CartoDB.Positron") %>%

  # Heatwave layer
  addPolygons(
    fillColor = ~pal_hw(mean_hw_3day_97_5),
    weight = 1,
    color = "white",
    opacity = 1,
    fillOpacity = 0.7,
    label = labels_hw,
    group = "Heatwaves"
  ) %>%

  # SVI layer
  addPolygons(
    fillColor = ~pal_svi(RPL_THEMES),
    weight = 1,
    color = "white",
    opacity = 1,
    fillOpacity = 0.7,
    label = labels_svi,
    group = "SVI"
  ) %>%

  # Add legends
  addLegend("bottomright", pal = pal_hw, values = ~mean_hw_3day_97_5,
            title = "Avg. 3-Day Heatwaves", group = "Heatwaves") %>%

  addLegend("bottomright", pal = pal_svi, values = ~RPL_THEMES,
            title = "Social Vulnerability Index", group = "SVI") %>%

  # Layer controls
  addLayersControl(
    baseGroups = c("SVI", "Heatwaves"),
    options = layersControlOptions(collapsed = FALSE)
  )

```



## Notes on analysis 

Join to weekly heat exposure data using tract ID (FIPS)

Exploratory Data Analysis

4.1 Time trends Plot yearly average heat days per tract

Plot number of heat waves per year

4.2 Spatial patterns Map of average weekly heat days per tract

Map of SVI values per tract

Map showing both for visual overlap

Statistical Analysis

5.1 Correlation analysis Correlation between average weekly heat
exposure and SVI

Include Pearson or Spearman correlation and scatterplot

5.2 Group comparison Create SVI categories (quartiles or high/low)

Compare average heat exposure across SVI groups (boxplots, t-tests or
Kruskal-Wallis)


```{r session.info, echo=FALSE}
#-----session info-----

sessionInfo()
```
