---
title: "Downloading NOAA Data"
output: html_document
date: "2025-07-03"
---

```{r set.knitr.options, include= FALSE, message=FALSE, warning=FALSE}
#-----set knitr options-----

# clear environment
rm(list = ls())

# knitr options
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, echo = TRUE, message = FALSE)

# Load pacman, installing if needed and other packages as needed
if (!require("pacman")) { install.packages("pacman", repos = "http://cran.r-project.org") }

# Load other packages, installing as needed.
pacman::p_load(tidyverse, dplyr, httr, jsonlite, tigris, sf, gstat, daymetr)

```

# About SVI Data 

The Social Vulnerability Index (SVI), developed by the CDC/ATSDR, measures the resilience of communities when confronted by external stresses such as natural disasters or disease outbreaks. It is based on 16 U.S. Census variables grouped into four themes: socioeconomic status, household composition & disability, minority status & language, and housing type & transportation. The SVI is reported at the census tract level and ranges from 0 (least vulnerable) to 1 (most vulnerable), allowing for comparison across geographic areas.

In our analysis, we use the 2020 SVI to assess how extreme heat exposure correlates with social vulnerability across San Francisco census tracts from 2010 to 2020. By linking weekly heat metrics derived from Daymet data to the SVI scores of each tract, we examine whether communities with higher social vulnerability experience more frequent or intense extreme heat events. This helps us identify environmental justice patterns and prioritize areas for future public health interventions.


## Downloading Data

Since no specific API exists, we can directly download from the website:  https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html

You can select Year: 2020, Geography: California, Geography Type: Census Tracts. 

```{r}

# read in data 
svi_data <- read_csv("Climate Change/Data/CA_SVI.csv")
View(CA_SVI)

```



## Data Preparation

Join to weekly heat exposure data using tract ID


Exploratory Data Analysis
4.1 Time trends
Plot yearly average heat days per tract

Plot number of heat waves per year

4.2 Spatial patterns
Map of average weekly heat days per tract

Map of SVI values per tract

Map showing both for visual overlap


Statistical Analysis
5.1 Correlation analysis
Correlation between average weekly heat exposure and SVI

Include Pearson or Spearman correlation and scatterplot

5.2 Group comparison
Create SVI categories (quartiles or high/low)

Compare average heat exposure across SVI groups (boxplots, t-tests or Kruskal-Wallis)




```{r session.info, echo=FALSE}
#-----session info-----

sessionInfo()
```
