---
title: "Heatwave Analysis"
output: html_document
date: "2025-07-17"
---

```{r set.knitr.options, include= FALSE, message=FALSE, warning=FALSE}
#-----set knitr options-----

# clear environment
rm(list = ls())

# knitr options
suppressMessages(library(knitr))
opts_chunk$set(tidy = FALSE, echo = TRUE, message = FALSE)

# Load pacman, installing if needed and other packages as needed
if (!require("pacman")) { install.packages("pacman", repos = "http://cran.r-project.org") }

# Load other packages, installing as needed.
pacman::p_load(tidyverse, scales)


# Set working directory 
setwd("~/PINC 2025/Bilaoen-PSP-2025/Climate Change Data")


```

# Heatwaves

According to the CDC (Centers for Disease Control and Prevention), a heatwave is a period of abnormally and uncomfortably hot and unusually humid weather, typically lasting two or more days. The definition can vary by location, but heatwaves are generally characterized by temperatures that are significantly higher than the average for a given area and time of year, and they can pose serious health risks, especially to older adults, children, and people with chronic conditions. For more information see: <https://www.cdc.gov/climate-health/php/resources/protect-yourself-from-the-dangers-of-extreme-heat.html>

Climate Change and Extreme Heat Events: <https://www.cdc.gov/climate-health/media/pdfs/ClimateChangeandExtremeHeatEvents.pdf>

Heatwave based on temperature and vapor pressure.

NCEI (National Centers for Environmental Information) calculates the 85th-percentile threshold for each MSA, both for minimum daily temperatures and for maximum daily temperatures. Using these data, EPA compared each city’s daily minimum apparent temperature records with that city’s corresponding 85th-percentile threshold. EPA identified a heat wave whenever two or more consecutive days exceeded the threshold, then quantified the following four metrics for each city and each year:

• Frequency: the number of distinct heat waves that occur every year.

• Duration: the length of each individual heat wave, in days. These data can be aggregated to find the average duration of individual heat waves over a time period such as a year or a decade.

• Season length: the number of days from the first day of the first heat wave of the year to the last day of the last heat wave, including the first and last days in the count.

• Intensity: how hot the temperature is during a heat wave, compared with the corresponding city-specific threshold. For example, if a city has an 85th-percentile threshold of 95°F, and the average of the daily minimum apparent temperatures during a three-day heat wave was 98°F, the intensity would be recorded as 3°F above the threshold.

Hence, health warnings about extreme heat are often based on NWS’s Heat Index, which is similar to apparent temperature in that it combines temperature and humidity (albeit with a different formula). This indicator specifically uses daily minimum temperature because studies show that mortality is more closely related to elevated daily minimum temperature than to daily maximum or daily mean temperature (Habeeb et al., 2015; Sarofim et al., 2016), as warm nighttime temperatures prevent the body from cooling off after a hot day. For each MSA, NCEI calculated daily maximum and minimum apparent temperature for each day based on hourly temperature and humidity measurements.

NCEI derived apparent temperature using the following equation:

AT = -1.3 + 0.92T + 2.2e

where AT is the apparent temperature (°C), T is ambient air temperature (°C), and e is water vapor pressure (kilopascals). This equation was established by Steadman (1984).

Based on Huang et al. (2021) we will examine the 97.5th percentile of apparent temperature, based on the minimum temperature. We will use the NCEI equation for apparent temperature, which does differ from Huang et al.'s (2021) equation that uses dew point instead.

Definitions:

1.  AT \> 85th percentile for 2+ days

2.  AT \> 85th percentile for 3+ days

3.  AT \> 97.5th percentile for 2+ days

4.  AT \> 97.5th percentile for 3+ days

```{r load.data, echo= FALSE}

# Step 0: Load downloaded and pre-processed clean daymet data

sf_daymet <- read_csv("sf_daymet_clean.csv")

```

Water vapor pressure vp Pa Water vapor pressure in pascals. Daily average partial pressure of water vapor.

We need to transform the vp variable from pascals to kilopascals. We divide the pressure value by 1000.

```{r apparent.temperature, echo = FALSE}
#  unique(sf_daymet$measurement)
# [1] "dayl" "prcp" "srad" "swe"  "tmax" "tmin" "vp" 

#Calculate tmean
# temp_data <- sf_daymet %>%
#   filter(measurement %in% c("tmax", "tmin")) %>%
#   pivot_wider(names_from = measurement, values_from = value) %>%
#   mutate(tmean = (tmax + tmin) / 2)


# Step 1: pivot data to wide format 
temp_data <- sf_daymet %>%
  filter(measurement %in% c("tmax", "tmin", "vp")) %>%
  pivot_wider(names_from = measurement, values_from = value) 

# kilopascals 
temp_data <- temp_data %>% 
              mutate(vp_kp = vp/1000)

# add AT 

# AT = -1.3 + 0.92T + 2.2e

temp_data <- temp_data %>% 
              mutate(AT = (-1.3 + 0.92 * (tmin) + 2.2 * (vp_kp)) )

# Reshape to long format
long_temp <- temp_data %>%
  select(site, year, yday, tmax, tmin, vp, vp_kp, AT) %>%
  pivot_longer(cols = c(tmax, tmin, vp, vp_kp, AT),
               names_to = "measurement",
               values_to = "value")

# Calculate 97.5th percentile per site & measurement

# tenyear_percentiles <- long_temp %>%
#   filter(year >= 2000 & year <= 2010) %>%     # Optional: first 10 years of data 
#   group_by(site, measurement) %>%
#   mutate(p97_5 = quantile(value, probs = 0.975, na.rm = TRUE), # percentiles are VERY similar when looking at 2000s
#             p85 = quantile(value, probs = 0.85, na.rm = TRUE))
# 


percentiles <- long_temp %>%
  filter(year >= 2000 & year <= 2020) %>%     # using 2000-2020 data in analysis 
  group_by(site, measurement) %>%
  mutate(p97_5 = quantile(value, probs = 0.975, na.rm = TRUE),
            p85 = quantile(value, probs = 0.85, na.rm = TRUE))


# filtering to only keep the AT data 
percentiles <- percentiles %>% filter(measurement == "AT")


```

How many "hot days" do we have per year per site (by year)?

```{r hot.days, echo= FALSE}

# Add threshold and flag hot days
long_temp_flagged <- percentiles %>%
  mutate(hot_day_85 = value >= p85,
         hot_day_97_5 = value >= p97_5)

# count how many days are flagged as hot days 
hot_days_summary <- long_temp_flagged %>%
  group_by(site, year) %>%
  summarise(
    n_hot_day_85 = sum(hot_day_85, na.rm = TRUE),
    n_hot_day_97_5 = sum(hot_day_97_5, na.rm = TRUE),
    .groups = "drop"
  )

# View first six rows of summary
head(hot_days_summary)


# visualizing


# ggplot(hot_days_summary, aes(x = year)) +
#   geom_line(aes(y = n_hot_day_85, color = "85th percentile")) +
#   geom_line(aes(y = n_hot_day_97_5, color = "97.5th percentile")) +
#   facet_wrap(~ site, scales = "free_y") +
#   labs(
#     title = "Number of Hot Days per Year by Site",
#     x = "Year", y = "Number of Hot Days",
#     color = "Threshold"
#   ) +
#   theme_minimal() + 
#  theme(plot.title = element_text(hjust = 0.5))




# Reshape for plotting
# hot_days_summary_long <- hot_days_summary %>%
#   pivot_longer(cols = starts_with("n_hot_day"),
#                names_to = "threshold", values_to = "count")
# 
# ggplot(hot_days_summary_long, aes(x = factor(year), y = count, fill = threshold)) +
#   geom_col(position = "dodge") +
#   facet_wrap(~ site, scales = "free_y") +
#   labs(
#     title = "Number of Hot Days per Year by Threshold",
#     x = "Year", y = "Count",
#     fill = "Threshold"
#   ) +
#   theme_minimal()  + 
#  theme(plot.title = element_text(hjust = 0.5))



# averages across city - average number of hot days per year 
hot_days_avg <- hot_days_summary %>%
  group_by(year) %>%
  summarise(
    mean_hot_day_85 = mean(n_hot_day_85, na.rm = TRUE),
    mean_hot_day_97_5 = mean(n_hot_day_97_5, na.rm = TRUE),
    .groups = "drop"
  ) 


ggplot(hot_days_avg, aes(x = year)) +
  geom_line(aes(y = mean_hot_day_85, color = "85th percentile")) +
  geom_line(aes(y = mean_hot_day_97_5, color = "97.5th percentile")) +
  labs(
    title = "Average Number of Hot Days per Year Across All SF",
    x = "Year", y = "Average Hot Days",
    color = "Threshold"
  ) +
  theme_minimal()  + 
 theme(plot.title = element_text(hjust = 0.5))


```

Now lets calculate the heat waves! We will use the four definitions we described above.

```{r heatwave.calculation, eval = FALSE}


long_temp_flagged_new <- long_temp_flagged %>%
  group_by(site, measurement, year) %>%
  arrange(yday, .by_group = TRUE) %>%
  mutate(
    # Streaks of 2+ for hot_day_85
    streak_85_2 = {
      r <- rle(hot_day_85)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 2
      ))
    },
    # Streaks of 3+ for hot_day_85
    streak_85_3 = {
      r <- rle(hot_day_85)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 3
      ))
    },
    # Streaks of 2+ for hot_day_97_5
    streak_97_5_2 = {
      r <- rle(hot_day_97_5)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 2
      ))
    },
    # Streaks of 3+ for hot_day_97_5
    streak_97_5_3 = {
      r <- rle(hot_day_97_5)
      inverse.rle(list(
        lengths = r$lengths,
        values = r$values & r$lengths >= 3
      ))
    }
  ) %>%
  ungroup()

# Preview the result
head(long_temp_flagged_new)

```

```{r heatwave.counting, echo = FALSE}

# Helper function to count qualifying streaks
count_streaks <- function(x, min_length) {
  r <- rle(x)
  sum(r$values & r$lengths >= min_length)
}

# Main heat wave summary
heat_wave_summary <- long_temp_flagged %>%
  filter(measurement == "AT") %>%
  group_by(site, year) %>%
  arrange(yday, .by_group = TRUE) %>%  # ensure correct order
  summarise(
    hw_2day_85   = count_streaks(hot_day_85, 2),
    hw_3day_85   = count_streaks(hot_day_85, 3),
    hw_2day_97_5 = count_streaks(hot_day_97_5, 2),
    hw_3day_97_5 = count_streaks(hot_day_97_5, 3),
    .groups = "drop"
  )

# save a copy of the heat wave data 

# write.csv(heat_wave_summary, "Data/heat_wave_summary.csv", row.names = FALSE)

  
```

# Answering Research Question 1: What are the temporal patterns of extreme heat events in San Francisco from 2000–2020?

Each team member can work on one of the heat wave definitions to answer the question above.

## Heatwaves: 2+ days at 85th percentile

```{r}
summary_hw_2day_85 <-   heat_wave_summary %>% 
    group_by(year) %>% 
    summarise(avg_hw_2day_85 = mean(hw_2day_85, na.rm = TRUE),
              .groups = "drop")
  
summary_hw_2day_85 %>%  ggplot(aes (x = year, y = avg_hw_2day_85)) + 
                                  geom_col(stat = "identity") + 
                                  geom_smooth(method = "loess")

# Create log-transformed variable
summary_hw_2day_85$log_avg_hw <- log(summary_hw_2day_85$avg_hw_2day_85)

# Fit linear model
model_log_hw_2day_85 <- lm(log_avg_hw ~ year, data = summary_hw_2day_85)

summary(model_log_hw_2day_85)

# interpret coefficient

exp(coef(model_log_hw_2day_85)["year"]) - 1


exp(0.023) - 1 # 0.0233 is 2.33% increase per year



# linear model 

model_lm_hw_2day_85 <- lm(avg_hw_2day_85 ~ year, data = summary_hw_2day_85)

summary(model_lm_hw_2day_85)

# year           0.23996    0.09163   2.619 
# p value   0.0169 * IS statistically significant 

# Each additional year is associated with an increase of ~0.24 heat wave events per year

# A linear regression model showed a statistically significant increase in the number of 2-day heat wave events above the 85th percentile in San Francisco between 2000 and 2020 (β = 0.24, p = 0.017). This corresponds to an average rise of about one additional 2-day heatwave every 4 years, highlighting a clear upward trend in short-duration heat extremes.



```

## Heatwaves: 3+ days at 85th percentile

```{r}

  summary_hw_3day_85 <-   heat_wave_summary %>% 
    group_by(year) %>% 
    summarise(avg_hw_3day_85 = mean(hw_3day_85, na.rm = TRUE),
              .groups = "drop")
  
  summary_hw_3day_85 %>%  ggplot(aes (x = year, y = avg_hw_3day_85)) + geom_col(stat = "identity") + geom_smooth(method = "loess")
  

# Create log-transformed variable
summary_hw_3day_85$log_avg_hw <- log(summary_hw_3day_85$avg_hw_3day_85)

# Fit linear model
model_log_hw_3day_85 <- lm(log_avg_hw ~ year, data = summary_hw_3day_85)
summary(model_log_hw_3day_85)

# interpret coefficient

exp(coef(model_log_hw_3day_85)["year"]) - 1


exp(0.03251644) - 1 # 0.03305 is 3.31% increase per year


# linear model 

model_lm_hw_3day_85 <- lm(avg_hw_3day_85 ~ year, data = summary_hw_3day_85)

summary(model_lm_hw_3day_85)
  
# year           0.17443    0.08894   1.961   

# p value 0.0647 is NOT statistically significant 

# Each additional year is associated with an increase of ~0.17 heat wave events per year

# A linear model examining the average number of 3-day heat wave events above the 85th percentile from 2000–2020 showed a marginally significant increasing trend (p = 0.065). The estimated increase was approximately 0.17 events per year, indicating a possible rise in multi-day extreme heat events, though the trend does not reach conventional statistical significance.


```

## Heatwaves: 2+ days at 97.5th percentile

```{r}

summary_hw_2day_97_5 <-   heat_wave_summary %>% 
    group_by(year) %>% 
    summarise(avg_hw_2day_97_5 = mean(hw_2day_97_5, na.rm = TRUE),
              .groups = "drop")
  
summary_hw_2day_97_5 %>%  ggplot(aes (x = year, y = avg_hw_2day_97_5)) + geom_col(stat = "identity") + geom_smooth(method = "loess")
  


# linear model 

model_lm_hw_2day_97_5 <- lm(avg_hw_2day_97_5 ~ year, data = summary_hw_2day_97_5)

summary(model_lm_hw_2day_97_5)

# year           0.12739    0.06943   1.835  

# p value 0.0822 NOT statistically significant 

# Each additional year is associated with an increase of ~0.13 2-day heatwave events per year

# Between 2000 and 2020, the average number of 2-day heat wave events above the 97.5th percentile increased by approximately 0.13 events per year — or roughly 1 additional event every 8 years.



```

## Heatwaves: 3+ days at 97.5th percentile

```{r}

  summary_hw_3day_97_5 <-   heat_wave_summary %>% 
    group_by(year) %>% 
    summarise(avg_hw_3day_97_5 = mean(hw_3day_97_5, na.rm = TRUE),
              .groups = "drop")
  
  summary_hw_3day_97_5 %>%  ggplot(aes (x = year, y = avg_hw_3day_97_5)) + geom_col(stat = "identity") + geom_smooth(method = "loess")
  
  


# linear model 

model_lm_hw_3day_97_5 <- lm(avg_hw_3day_97_5 ~ year, data = summary_hw_3day_97_5)

summary(model_lm_hw_3day_97_5)

# year           0.06876    0.03092   2.224   
# Each additional year is associated with an increase of ~0.069 events per year.

# p value 0.0385 * IS statistically significant 
  
# From 2000 to 2020, the average number of 3-day heat wave events (above the 97.5th percentile) in San Francisco increased by approximately 0.069 events per year, a statistically significant upward trend (p = 0.0385).


```

```{r}

all_summary <-  heat_wave_summary %>% 
    group_by(year) %>% 
    summarise(avg_hw_2day_85 = mean(hw_2day_85, na.rm = TRUE),
              avg_hw_3day_85 = mean(hw_3day_85, na.rm = TRUE),
              avg_hw_2day_97_5 = mean(hw_2day_97_5, na.rm = TRUE),
              avg_hw_3day_97_5 = mean(hw_3day_97_5, na.rm = TRUE),
              .groups = "drop") 
  


library(tidyverse)

# Assuming your tibble is called all_summary
all_summary_long <- all_summary %>%
  pivot_longer(
    cols = starts_with("avg_hw"),
    names_to = "heat_metric",
    values_to = "value"
  )

# Now plot using ggplot with facet_wrap
ggplot(all_summary_long, aes(x = year, y = value)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ heat_metric, scales = "fixed") +
  labs(
    title = "Annual Average Heatwave Metrics",
    x = "Year",
    y = "Average Heatwaves Events",
    caption = "Faceted by heatwave definition"
  ) +
  theme_minimal()


ggplot(all_summary_long, aes(x = factor(year), y = value)) +
  geom_col(fill = "#b39dd0", alpha = 0.8) +
  geom_smooth(aes(group = 1), method = "loess", color = "#ae017e", se = FALSE) +
  facet_wrap(~ heat_metric, scales = "fixed") +
  labs(
    title = "Annual Average Heatwave Metrics",
    x = "Year",
    y = "Average Heatwave Events",
    caption = "Faceted by heatwave definition"
  ) +
  theme_minimal() + 
  theme(
    axis.text = element_text(size = 8, face = "bold"))

```






# Heatwave Change Over Time

We can also show the change in heatwave by definition over the last 20 years

First, we compared 2000 vs 2020. 

```{r}

# Calculate the increase from 2000 to 2020 

# Filter and reshape
heat_change_all <- heat_wave_summary %>%
  filter(year %in% c(2000, 2020)) %>%
  pivot_wider(
    id_cols = site,
    names_from = year,
    values_from = c(hw_2day_85, hw_3day_85, hw_2day_97_5, hw_3day_97_5),
    names_sep = "_"
  ) %>%
  mutate(
    hw_2day_85_diff     = hw_2day_85_2020 - hw_2day_85_2000,
    hw_3day_85_diff     = hw_3day_85_2020 - hw_3day_85_2000,
    hw_2day_97_5_diff   = hw_2day_97_5_2020 - hw_2day_97_5_2000,
    hw_3day_97_5_diff   = hw_3day_97_5_2020 - hw_3day_97_5_2000
  ) %>%
  select(site, ends_with("_diff"))



heat_change_all <- heat_change_all %>%
  mutate(site = as.character(sprintf("%011.0f", site)))


sf_neighborhoods <- read_csv("Analysis_Neighborhoods.csv")


# Join with EHE + Neighborhood Data

heat_change_sf <- sf_neighborhoods %>%
  left_join(heat_change_all, by = c("geoid" = "site"))

# add geometry 
# Convert to sf using WKT from 'the_geom' column
neighborhood_heat_change_sf <- heat_change_sf %>%
  st_as_sf(wkt = "the_geom", crs = 4326) 


# remove The Farallones

neighborhood_heat_change_sf <- neighborhood_heat_change_sf %>%  filter(geoid != "06075980401")



# Heatwave change in SF  
ggplot(neighborhood_heat_change_sf) +
  geom_sf(aes(fill = hw_2day_97_5_diff), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "magma", na.value = "grey90") +
  labs(
    title = "Change in Heatwave Events by Census Tract (2000 - 2020)",
    fill = "Average Annual Heatwaves 2day 97.5th"  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )




# Heatwave change in SF  


tmap_mode("plot")

tm_shape(neighborhood_heat_change_sf) +
  tm_fill("hw_2day_97_5_diff",
          palette = "magma",
          title = "Change in 2-Day 97.5th perecentile  (2000–2020)",
          midpoint = 0,
          colorNA = "lightgray") +
  tm_borders() +
  tm_layout(
    title = "Change in Heatwave Events by Census Tract",
    frame = FALSE,
    legend.outside = TRUE
  )


# you can also make faceted versions of the map 

neighborhood_heat_change_long <- neighborhood_heat_change_sf %>%
  pivot_longer(
    cols = starts_with("hw_"),
    names_to = "definition",
    values_to = "change"
  ) %>%
  mutate(definition = recode(definition,
    "hw_2day_85_diff"   = "2-Day ≥ 85th",
    "hw_3day_85_diff"   = "3-Day ≥ 85th",
    "hw_2day_97_5_diff" = "2-Day ≥ 97.5th",
    "hw_3day_97_5_diff" = "3-Day ≥ 97.5th"
  ))



# using GGPLOT 

ggplot(neighborhood_heat_change_long) +
  geom_sf(aes(fill = change), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "magma", na.value = "grey90") +
  facet_wrap(~ definition, ncol = 2) +
  labs(
    title = "Change in Heatwave Events by Census Tract (2000–2020)",
    fill = "Change in Events"
  ) +
  theme_void() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )



# Using TAP

tmap_mode("plot")

tm_shape(neighborhood_heat_change_long) +
  tm_fill("change",
          palette = "magma",
          title = "Change (2000–2020)",
          midpoint = 0,
          colorNA = "lightgray") +
  tm_borders(lwd = 0.2) +
  tm_facets(by = "definition", ncol = 2) +
  tm_layout(
    title = "Change in Heatwave Events by Census Tract",
    legend.outside = TRUE,
    frame = FALSE
  )

```


We also looked at the trend over the 20 years (considers change over time)

```{r}
library(dplyr)
library(purrr)
library(broom)
library(tidyr)

# Reshape to long format so we can model all definitions at once
heat_trends_long <- heat_wave_summary %>%
  pivot_longer(
    cols = starts_with("hw_"),
    names_to = "definition",
    values_to = "count"
  )

# Fit linear models per site and definition, then extract slope
heat_trend_slopes <- heat_trends_long %>%
  group_by(site, definition) %>%
  summarise(
    slope = coef(lm(count ~ year))[2],
    .groups = "drop"
  )


heat_trend_slopes <- heat_trend_slopes %>%
  mutate(site = as.character(sprintf("%011.0f", site)))


neighborhood_trend <- neighborhood_sf %>%
  left_join(heat_trend_slopes, by = c("geoid" = "site"))

neighborhood_trend <- neighborhood_trend %>%
  mutate(definition = recode(definition,
    "hw_2day_85"   = "2-Day ≥ 85th %tile",
    "hw_3day_85"   = "3-Day ≥ 85th %tile",
    "hw_2day_97_5" = "2-Day ≥ 97.5th %tile",
    "hw_3day_97_5" = "3-Day ≥ 97.5th %tile"
  ))


# using GGPLOT 

neighborhood_trend <- neighborhood_trend %>%
  mutate(definition = recode(definition,
    "hw_2day_85"   = "2-Day ≥ 85th",
    "hw_3day_85"   = "3-Day ≥ 85th",
    "hw_2day_97_5" = "2-Day ≥ 97.5th",
    "hw_3day_97_5" = "3-Day ≥ 97.5th"
  ))

ggplot(neighborhood_trend) +
  geom_sf(aes(fill = slope), color = "white", size = 0.2) +
  scale_fill_distiller(
    palette = "RdYlBu",
    direction = -1,
    na.value = "grey90",
    name = "Slope (Events/Year)"
  ) +
  facet_wrap(~ definition, ncol = 2) +
  labs(
    title = "Heatwave Event Trends by Census Tract (2000–2020)"
  ) +
  theme_void() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom"
  )



# Using TMAP
tmap_mode("plot")

tm_shape(neighborhood_trend) +
  tm_fill("slope",
          palette = "-RdYlBu",
          midpoint = 0,
          title = "Slope (Events per Year)",
          colorNA = "lightgray") +
  tm_borders(lwd = 0.2) +
  tm_facets(by = "definition", ncol = 2, free.scales = FALSE) +
  tm_layout(
    title = "Heatwave Event Trends by Census Tract (2000–2020)",
    legend.outside = TRUE,
    frame = FALSE
  )

#plotting change over time 

```



